Comparing variability across fMRI analysis software packages and numerical errors

### 2021-12-17 paper/paper.tex
Rewrote intro

\maketitle
\begin{abstract}
  Variability has been broadly observed in functional MRI analyses across both
  differences in analytic tools and numerical instabilities. However,
  the relationship between numerical (within-tool) and between-tool variabilities for
  given analysis is unclear. We extended a previous comparison
  of fMRI analysis software libraries (namely FSL, AFNI, and SPM) and relate the observed differences to the numerical
  variability observed using Monte-Carlo arithmetic. We found that
  between-tool variability was consistently larger than numerical variability in group-level analyses,
  whereas numerical variability approached between-tool variability in some subject-level analyses.
  Interestingly, between-tool variability and
  machine error appeared moderately correlated ($r > 0.5$) in all tested conditions.
  Finally, we found that numerical variability had effects of similar magnitude to between-tool
  variability when errors were introduced at the precision of 17 bits --- a
  precision level found in commonly-used libraries \gk{What does this mean? They only have precision of 17, or they use data that is 17 bits?}. \yohan{agree with Greg, it is unclear and a strong assertion!} \camille{It is not entirely clear to me how this relates to the previous conclusions that BT variability was consistently larger than machine error} Our findings motivate the
  continued investigation of numerical instability in neuroimaging, and position it
  as a possible proxy for studying arbitrary between-tool variations.
\end{abstract}

@@

Software variability has been highlighted as an important source of error
in fMRI analysis, originating from mainly two factors: between-tool
variability denotes differences in the outcomes of different
software toolboxes --- such as FSL, AFNI and SPM --- when analyzing the same dataset
using similar methods \gk{this sentence says "two factors" but then only mentioned between tool. Also, it is too long}. \yohan{What do you mean by methods? Numerical schemes? If so, are we sure that theses tools are using the same numerical schemes?} \camille{I also had highlighed the 2 here, I think we could use "multiple factors" --> in practice we could also add different types of within-software variability (in the form of changing algorithms and/or parameters for a given step)} Numerical variability, in contrast, denotes
differences in outcomes of the same analytical toolbox
executed in different environments – such as different operating systems or
different hardware platforms \gk{you mention these two examples of numerical variability, but you use neither in this paper. Why not mention the one you use?}. This manuscript investigates the extent to
which both types of variability relate to each other.

Numerical variability stems from errors introduced in floating point
computations by \yohanmod{numerical} rounding, cancellation, and absorption~\cite{muller2018handbook} \gk{only? What about poor conditioning + random effects + noise in data?}. In
practice, the occurrence of such errors is affected by variations in
hardware, software dependencies, compilation, parallelization, and most
likely other factors \gk{what does "most likely other factors" mean?}. In the best case, the impact of such errors and the induced
variations is in the range of \yohanmod{1} \yohan{1/2 ulp for round-to-nearest} unit in the last place (ulp) --- also known
as machine error, the difference between two successive floating-point
numbers. However, when computations depend on imprecise libraries,
numerical error may vary by several ulps. For instance, the \yohan{maximal} numerical error
of the complementary error function (erfc) is larger than 3 ulps in several
mathematical libraries~\cite{zimmermann:hal-03141101}, implying that an update of this function could introduce
numerical \yohanmod{perturbations} \yohan{differences} larger than 3 ulps. Due to numerical instability,
machine error may amplify during its propagation through analyses \yohan{do you mean calculations?},
resulting in considerable loss of precision in final results. In fMRI analysis,
numerical variability produced by Linux updates has been shown to result in Dice
coefficients ranging from 0.0 to 1.0 for ICA components produced by FSL
MELODIC~\cite{Glatard2015} \gk{there isn't really context for what this result means, yet. Is this a lot? Dice hasn't been explained. Also, how are ICA components (weighted vectors) compared via Dice (a measure for binary image/vector comparison)? }.

Between-tool variability is not a surprising concept: different programs
executed on the same data may obviously produce different results \gk{even when doing conceptually the same thing? Why? Why is this obvious?}. However,
in fMRI analysis, programs all aim to estimate brain activity from measured
variations in the BOLD signal, which is not expected to be measurably
impacted by computational artifacts \gk{Why wouldn't it be? Especially if you state that it isn't surprising for tools to lead to different results, how come you expect fMRI to not suffer from this?}. \yohan{+1} Therefore, the observation reported
in~\cite{bowring2019exploring} that the three main fMRI analysis toolboxes
may result in marked analytical differences (Dice coefficients ranging from
0.000 to 0.684 \camille{Apologies I don't have the numbers on the top of my head but can I confirm taht those were taken from the erratum that was published after the paper?} for thresholded activation maps), has raised substantial
concern. In a follow-up study~\cite{bowring2021isolating}, the same authors
isolated the analytical components where the pipelines diverge,
highlighting a major impact of the signal model and suggesting that the
root-cause for between-tool variability depends on the analysis design and
task paradigm \gk{If you state this finding, you need to unpack what it means a little bit. What part of analysis design? By task paradigm do you essentially mean the expected effect-size for the task that individuals are performing within the scanner?} \camille{would be worth adding here that the identified main sources of variations were different across the 3 studies}. Further, the work in~\cite{Li2021.12.01.470790} found
moderate agreement between results produced by five independent fMRI
analysis pipelines and analyzed in details the components that led to these
variations \gk{This sentence feels inconsistent and counter to the previous, and also doesn't tell us their findings; what components led to variations?}. Between-tool variability is supposedly an important determinant
of the substantial analytical variability recently observed in
fMRI~\cite{botvinik2020variability} \gk{arguably, one of the larger issues that the NARPS result showed us was that people construct and test ``equivalent'' hypotheses differently, so despite highly similar data they will arrive at different conclusions}.

We investigate the relationship between numerical and between-tool
variability through two main questions: (1) what is the relative magnitude of
within-tool (numerical) variability and between-tool variability, and (2) is there an association between within- and between-tool variability. The first question determines the extent to which each variability
source may impact current fMRI results, and which one should be addressed or leveraged
in priority. The second question may provide deeper insights on the origins
of software variability. Conceptually, both types of variability may
reflect a single solution space shaped by local minima and
ill-conditioning \gk{this feels a bit clumsily worded, but I don't immediately have a rewrite that comes to mind }, which could be explored through numerical or software
perturbations.

@@

We evaluated WT and BT variability for thresholded as well as unthresholded
group-level and subject-level t-statistics maps by computing the
standard deviation of t-statistic maps across pipelines or Fuzzy Libmath samples, respectively. For BT, we computed the standard deviation
across a given pair of tools \camille{standard dev with 2 samples ? --> cf. discussion on mattermost).
For WT, we computed the standard deviation across the three
Fuzzy Libmath samples of a given tool. We computed WT variability for a pair of tools (A, B) as follows:
\begin{equation}
  \sigma_{WT(A,B)}^2 = \sigma_{WT(A)}^2 + \sigma_{WT(B)}^2,
  \label{eq:wt-pair}
\end{equation}
where $\sigma_{WT(.)}$ is the numerical variability for a given tool
and $\sigma_{WT(., .)}$ is the numerical variability for a pair of tools.

>>>>

\maketitle
\begin{abstract}
  Variability has been broadly observed in functional MRI analyses across both
  differences in analytic tools and numerical instabilities. However,
  the relationship between numerical (within-tool) and between-tool variabilities for
  a given analysis is unclear. We extended a previous comparison
  of fMRI analysis software libraries (namely FSL, AFNI, and SPM) and relate the observed differences to the numerical
  variability observed using Monte-Carlo arithmetic. We found that
  between-tool variability was consistently larger than numerical variability in group-level analyses,
  whereas numerical variability approached between-tool variability in some subject-level analyses.
  Interestingly, between-tool variability and
  machine error appeared moderately correlated ($r > 0.5$) in all tested conditions.
  Finally, we found that numerical variability had effects of similar magnitude to between-tool
  variability when numerical errors are introduced at the precision of 17 bits --- a
  precision level found in commonly-used libraries \gk{What does this mean? They only have precision of 17, or they use data that is 17 bits?}.
   \yohan{agree with Greg, it is unclear and a strong assertion!}
   \camille{It is not entirely clear to me how this relates to the previous conclusions that BT variability 
   was consistently larger than machine error} Our findings motivate the
  continued investigation of numerical instability in neuroimaging, and position it
  as a possible proxy for studying arbitrary between-tool variations.
\end{abstract}

@@

Recent explorations of the analytical flexibility in brain imaging across
tools, platforms, or teams, have demonstrated unexpected variability in
results, even when analyzing identical data~\cite{botvinik2020variability}. Possible explanations for
such discrepancies include methodological flexibility~\cite{carp2012plurality} and software
variability, the focus of this paper. A recommendation to address this
variability is to adopt a ``multiverse" approach that analyzes the same
dataset multiple times in different software environments, and ultimately
conclude from the resulting set of outcomes. However, the range of
analytical conditions to be included in such multiverse analyses remains
poorly defined, both because of the boundless set of tools and
configurations, and that the precise causes of software variability remain
unclear. This lack of clarity is in part because fMRI analyses depend on
complex software stacks that leverage low-level libraries provided by the
operating system (e.g.,mathematical functions), general scientific
computing methods (e.g., optimization toolboxes), and specific fMRI
analysis tools (e.g., spatial image normalization methods). At each level,
conceptual and implementation differences across experiments may each
create substantial variability in the analysis outcomes.

As a result of these factors, the variability resulting from the use of
different fMRI analysis tools implementing similar analytical approaches
(between-tool variability) can reach worrying magnitudes. For instance, the
study in~\cite{bowring2019exploring} compared the results produced by the
three main fMRI analysis toolboxes, namely SPM, AFNI and FSL, using similar
pipelines. It reported limited similarity between the activation clusters
produced by these tools, measured by Dice coefficients ranging from 0.0 to
0.684 \tristan{check values in erratum}, where 0 indicates non-overlapping
clusters and 1 means identical clusters. More recently, the work
in~\cite{Li2021.12.01.470790} also showed a low similarity between the
results produced by five different tools (ABCD, CCS, CPAC, DPARSF, fMRIPrep
\tristan{Ali, could you add refs to these tools from Li et al 2021? Also
for SPM, FSL and AFNI in the previous sentence}) and identified the main
factors contributing to these differences. The magnitude of the highlighted
differences suggest that between-tool variability may be playing a critical
role in the reproducibility of fMRI analysis overall.

The variability resulting from differences in lower-level software
libraries (within-tool variability) has also been quantified in fMRI. The
study in~\cite{Glatard2015} mentions a low similarity between the activation clusters
produced by FSL using different versions of the GNU/Linux system, measured
by Dice coefficients ranging from 0.0 to 1.0, covering the full spectrum of
possible similarities. This variability resulted from updates in the GNU
mathematical library and can be properly simulated by introducing small
numerical perturbations on the results returned by mathematical functions~\cite{salari2021accurate}.
Here again, these observations could partly explain the differences
reported in~\cite{botvinik2020variability}.

The relationship between within- and between-tool variability are poorly
understood, but could play an important role in the construction of
multiverse study environments. Importantly, if associations exist between
these two types of variability, they may both originate to some degree from
the inherent instability of brain activity estimation from BOLD signal
variations, and shed light on the numerical confidence of results. Under
this hypothesis, small perturbations introduced by low-level software
updates could trigger effects correlated with those created by tool
variations.

This paper investigates the relationship between numerical stability -- used
as a proxy for within-tool variability -- and between-tool variability
through two main questions:
\begin{enumerate}
\item what is the relative magnitude of within-tool variability and between-tool variability, and
\item is there an association between within- and between-tool variability.
\end{enumerate}

We address these two questions by reproducing the study in~\cite{bowring2019exploring} and
extending it with the addition of numerical perturbations of controlled
magnitude.

@@

We evaluated WT and BT variability for thresholded as well as unthresholded
group-level and subject-level t-statistics maps by computing the
standard deviation of t-statistic maps across pipelines or Fuzzy Libmath samples, respectively. For BT, we computed the standard deviation
across a given pair of tools \camille{standard dev with 2 samples ? cf. discussion on mattermost}.
For WT, we computed the standard deviation across the three
Fuzzy Libmath samples of a given tool. We computed WT variability for a pair of tools (A, B) as follows:
\begin{equation}
  \sigma_{WT(A,B)}^2 = \sigma_{WT(A)}^2 + \sigma_{WT(B)}^2,
  \label{eq:wt-pair}
\end{equation}
where $\sigma_{WT(.)}$ is the numerical variability for a given tool
and $\sigma_{WT(., .)}$ is the numerical variability for a pair of tools.
